---
title: "Week 1:  Setting Up"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Week 1:  Setting Up}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(datascience.curriculum)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

## Introduction

This course will focus on what I believe are the essential skills necessary for a laboratory-based scientist to use R and Rstudio in an effective way.  There are many other detailed resources out there for you.  

* [Software carpentry](https://software-carpentry.org/lessons/previous/) has very good tutorials for R, python, git and working with the command line.  
* [R for data science](https://r4ds.had.co.nz/) is a book written by Hadley Wickham who wrote much of the software we use in R
* [Modern dive](https://moderndive.com/) is focused on modeling if you are interested in more clinical data sets 

These and many others are good resources if you want to extend your knowledge beyond what we cover here.    

Learning R at times will seem like learning a foreign language.  You will make errors.  **Always remember that you will never be the first person to encounter an error.  Learn how to google the error message and 99% of the time you will find your answer in the first 3 hits**.  This is how you learn. 

## R and Rstudio

R is a statistical programming language that runs as a computer program.  Rstudio is known as an integrated development environment or IDE.  It is the standard way humans interact with the R program.  

Rstudio provides two main ways of interacting with R:  directly via the R Console and via R scripts.

* R console:  you type in commands and output is usually returned to the console.  The commands are logged in a history file if you really need them, but this is inconvenient.  A better option is to use scripts.  
  
  **Warning and Error messages will appear in the console.  Always pay attention to what they are saying.** 

* R scripts:  these are text files, usually in the form of <filename>.R.  When you run 1 or more lines, Rstudio feeds them into the console and R follows the commands.  This is good because you always have a plain text record of what you did.  So when you build up a multi-step analysis it is easy to see how you got the end product.  Better than running the whole file line by line is "sourcing" the file.  You can do this by typing Ctrl+Shift+S or clicking the "Source" button in the upper right corner of the source pane (usually top-left) in Rstudio.  Why is this better?  Let's say you left an error in your script.  Maybe the presence of this error changes the final output you are trying to produce.  If you run all of the lines, R will notify you of the error and then keep going.  Easy to miss if this is the only error.  If you source the file, it will stop completely on errors so you don't end up with errors you aren't aware of.

  **You always want your programs to error in obvious ways so you don't carry on with incorrect data.**

Rstudio provides a number of standard panes that display useful information.  We will review these in class or you can explore them on your own.

## Setting up your system

I recommend this approach to setting up your R installation.  In comparison to other methods, the blaseRtemplates method provides greater transparency and reproducibility and an improved user experience (faster, fewer confusing messages).  But your mileage may vary.

First, install the blaseRtemplates package using whatever package installer you currently use. Assuming the base install function, run:

```{r}
install.packages('blaseRtemplates', repos = c('https://blaserlab.r-universe.dev'))
```

Next, simply run the following command:

```{r}
library("blaseRtemplates")
establish_new_bt(cache_path = "<some_directory>/r_4_2_cache", project_path = "<some_directory>/projects")
```

What will this do?

-   It will create directories where indicated, write a standard .Rprofile document, and modify the user .Renviron file to configure R properly.\
-   It will not overwrite existing directories and will fail when an existing installation is already present in the same location.\
-   The only modification outside of the provided directory is to the .Renviron file. If you have existing configurations in your .Renviron file, they will only be changed if they are conflicting. If this is problematic, then you should archive your .Renviron file before running. The new .Renviron file will cause R to use the new .Rprofile in the cache directory and skip existing user- and project-level .Rprofiles.

Briefly, when R starts a new session, it looks for a series of files to configure itself. This configuration includes environment variables (in the .Renviron file) and any desired R code (in the .Rprofile file). As included in this function, these configurations only affect user experience i.e. they should not affect the results produced by your code. By user experience I mean connection to the proper package libraries and repositories and some safeguards to prevent working outside of projects.

For more information on how R starts, see [this link.](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Startup.html "R startup").

### Directory Structure

    .
    |-- projects
    |   `-- baseproject
    `-- r_4_2_cache
        |-- library
        |-- logs
        |-- source
        |-- user_project
        |-- dependency_catalog.tsv
        |-- package_catalog.tsv
        `-- .Rprofile

First, note that two main subdirectories have been created in one parent directory. This is optional and can be specified by the function inputs, but for a single user system it is recommended to keep them together for the sake of organization.

-   projects: where your projects should be initialized. They can go anywhere but you should keep them here. The baseproject is created by the establish_new_bt function. If you close a project without moving to a new one or try to start a session outside of a project, it will bounce you here. This prevents undesired operations in e.g. the home directory or elsewhere.

-   r_4\_2_cache: This holds all of the inner workings of the blaseRtemplates system.

    -   library: The versioned/hashed package binary cache. Each package is stored under the following rubric: name \> version \> hash \> name \> contents. This structure is generated whenever blaseRtemplates installs a new package in any project.

    -   logs: collection of .Rhistory files. Rather than 1 endless file (default), blaseRtemplates will store history files for each session by user and date/time

    -   source: Package source files. Used by the installer, pak.

    -   user_project: A collection of symlinks to versioned/hashed packages in the library. This is identified as the first entry in .libPaths() so it is where R looks first for packages. Why do this? 1. It allows reproducibility. If I update a package for one project, it may break something in an older project. This allows each project to have its own set of packages. (The list of what versions are used is kept as a tsv file within the project, so it is completely portable. Anyone who wants to replicate your code just has to install that list of packages.) 2. It makes the system lightweight. Package code and more significantly, data, is stored in one location and accessible to many projects, according to version. 

    -   dependency_catalog.tsv: a list of package dependencies by package hash. Speeds up installation when linking packages from the cache to a project.

    -   package_catalog.tsv: the master catalog of packages available in the cache.

    -   .Rprofile: the Rprofile used on startup, as specified by the new .Renviron file

The library directory is initially populated using all packages available to R the local libraries.  Depending on what is in those libraries, and processing/network speeds, this may take some time.


## Projects

A project is a self-contained group of files for your R data analysis.  This should align conceptually with your experiments.  If you have a set of single cell data and imaging data related to the same scientific concept or paper then it can all go in the same project.

Projects automatically define their working directory to be the root of the projects.  That means you can reference any file in the root directory by typing its name like so: "file.R".  If that file was one directory down, it would be "directory/file.R". There is almost no good reason ever to change your working directory to something other than the project root.

You always want to be working in a project.

### Setting up a project

The best way to do this is through automated commands, rather than through the program menus.  This way the packages are set up in a consistent manner, every time.  Then when you have to revisit a project after months and months, it is easy to navigate.


1.  Choose the directory you want the project to live in.

2.  Run

```{r}
# create a project called rclass_example in your home directory
blaseRtemplates::initialize_project("~/rclass_example")
```

That function will create the skeletonized project, activate it and restart your session within it.

The following lines are taken directly from one of files created, "R/inititialization.R".  You should run these when you start up the project the first time:

```{r}
# after running the initialization command, run these lines once to set up
# your project in a consistent manner.
# private github repos can be made public via the web interface

# make a software license
usethis::use_mit_license("<your name here>")

# generate a readme file to explain your work
usethis::use_readme_md(open = FALSE)

# *** Only if developing a package ***
# uncomment and run to generate a news file to document updates.
# usethis::use_news_md()

# initialize git
usethis::use_git()

# initialize github
usethis::use_github(private = TRUE)


### Delete this file after initializing the project! ###
```

After you run this script, you should delete it.  See below for setting up git and github for the first time.

### Project organization

I keep all of my R scripts in a directory called "R".  If you initialized your project with ```blaseRtemplates::initialize_project()```, in addition to ```initialize.R```, you will have scripts labeled:  


* ```dependencies.R```:  use this for installing packages into your renv environment and attaching them to your session
* ```configs.R```:  use this for setting graphical and aesthetic parameters.  I also use this for resolving function name conflicts.
* ```git_commands.R```:  only necessary if you are using git (recommended).  The commands here will allow you to handle 99% of git interactions via R in a scripted way.

In addition to this, the main analysis scripts for the project will go in the R folder and may go in subdirectories if there are a lot of them.  Other file types, such as *.Rmd, would go in their own folder named Rmd/.  This arrangement is very flexible for simple data analysis projects; the rules become more strict when we build packages.

## Packages

Packages are chunks of code and/or data that people have written and made available to us to use. The main reason we use R is that there is a large library of useful packages for the biological sciences that we can build our work on, all of which are free to use. 

Packages are most often stored and distributed via standard repositories.  The most common are [CRAN](https://cran.r-project.org/), [R-universe](https://r-universe.dev/search/), and [Bioconductor](https://www.bioconductor.org/).  Packages can also be distributed as source code (text based files), usually via github. 

**What happens when we "install" packages?**

R downloads the compressed file and then unpacks it in a library directory on your computer or server.  IF you are installing from source code (most linux computers or github repositories), then the code needs to be turned into a set of binary instructions for the computer to use, a process called compiling.  This is what the Xcode (Mac) and RTools (Win) programs do.

Every time you start an R session, a set of commands is run that points R to one or more library directories.  This is how R finds the commands you ask it to use. Unfortunately, it can be difficult to find where these directories live, depending on your installation.  To help with this, I have developed a package management system, called blaseRtemplates, that we will use in this course (see below).

**How do I use a function?**

Once you have installed a package, its functions are available to be called using the package name followed by 2 colons such as ```dplyr::filter(mtcars, cyl == 6)```. But this is not always the best way to go.  

To make all of the functions from a package available, you need to run ```library("<package name>")```.  This will attach all of the functions from that package to your session.  Then you can just use the function name without the package prefix. 

Beware of attaching more packages than you actually need though because there could be "namespace conflicts".  This happens when the same term applies to two different functions from two different packages.  R picks one and goes with it without clearly indicating that it is doing so.  This is a common source of errors.

Later on we will see how to use the "conflicted" package to help warn us of those errors before they cause a problem. 

**Where do I look for help with R functions?**

A manual for each function in an R package is generated and can be viewed in the help panel in RStudio. Access these with ```?<object name>``` e.g. ```?mean```.  Vignettes may also be available.  Vignettes are long-form articles that show you how to use the functions in a package together, often with example data.  These are best accessed by navigating through the package menu.  Click on the package name and near the top you will see any available vignettes.  

### Installing packages

As mentioned above, blaseRtemplates-based projects use a user-project library that is linked to specific package versions in the main cache.  This way, each project you start can use its own set of packages without duplication of the actual software.  The best way to install packages in this format is this:

* Install a new package or get the latest version from a repository

``` r
blaseRtemplates::install_one_package(package = "<package_name>", how = "new_or_update")
```

* Get a fresh link to the newest version available in the cache

``` r
blaseRtemplates::install_one_package(package = "<package_name>", how = "link_from_cache")
```

* Get a specific version from the cache

``` r
blaseRtemplates::install_one_package(package = "<package_name>", how = "link_from_cache", which_version = "1.0.0.")
```

* Versioned package installations are supported for cran only.  If unavailable, the latest version will be installed.

``` r
blaseRtemplates::install_one_package(package = "<cran_package_name>", how = "new_or_update", which_version = "1.0.0.")

```

You can find a human-readable table with all of the packages in the cache at BLASERTEMPLATES_CACHE_ROOT/package_catalog.tsv.

You can find a human-readable table with all of the packages used by a given project in the library_catalogs directory of the project.  There is at least one file in there.  For projects collaborating through git, there will be one for each user.  These files indicate all available and actively used packages in the R code.  This file is meant to be tracked by git and so can provide version control and reproducibility for the packages used in your project.

### Get a new project library

You can use this function to replace the entire project library with a new one.  By default, the function will install the newest version of each package available available in the cache. 

``` r
blaseRtemplates::get_new_library()
```

Say, however, that you are collaborating on this project.  Another person may be using a different set of packages.  This can in some circumstances change output.  To adopt the package library used by a different individual, supply their project library catalog as the argument to ```get_new_library()```.

``` r
blaseRtemplates::get_new_library("library_catalogs/<file>.tsv")
```


## Extra credit:  Basic git operation for the single user

You want to try to use git and github.  There is a vast array of resources for you to read about git and troubleshoot issues.  Briefly, it is a program that tracks changes to text files.  Since all of the code you will write is text-based this is a great resource for you to keep track of your progress on your work.

Git tracks changes to files in a group of files known as a repository.  The way we are using it here, your project is the repository.  Github is simply a website that hosts a copy of your repository on the web.  This is how you will share your work with journal reviewers and editors and others.


```{r}
# make sure you have a github account
# https://github.com/join

# install git
## Windows ->  https://git-scm.com/download/win
## Mac     ->  https://git-scm.com/download/mac
## Linux   ->  https://git-scm.com/download/linux

# configure git in Rstudio
usethis::use_git_config(user.name = "YourName",
                        user.email = "your@mail.com")

# create a personal access token at the github website
# set the expiration date as desired (no expiration date)
# permissions should be set automatically.  Then run:
usethis::create_github_token()

# run this and enter your token at the prompt
blaseRtemplates::gitcreds_set()

# Run the following once per user to collaborate smoothly
blaseRtemplates::setup_git_collab()

```
Connecting/authenticating to github is the most difficult part of all of this.  The keychain program used by ```gitcreds_set()``` may not be available or working on all systems.  As a workaround, you can replace the line in your .Renviron file, GITHUB_PAT=gitcreds::gitcreds_get(use_cache = FALSE)$password, with GITHUB_PAT=<your token>.  There is a small security risk to this if someone were to gain access to your .Renviron file. So proceed with caution.

The necessary permissions should be set automatically by these commands.  By default, the token expires after 30 days, but you can override this.  

Assuming the commands above have run properly you should be ready to use git and github.  You can interact with git in at least 4 different ways:

* via scripted commands on ```R/git_commands.R```
* using the Git pane in Rstudio
* using the terminal
* using a third-party git client such as [git kraken](https://www.gitkraken.com/) or [sourcetree](https://www.sourcetreeapp.com/)

My recommended approach for new users is to use the scripted git commands to control what is happening and to use the Git pane to monitor the status of your repository.  Just remember to hit the refresh icon if things look strange in the Git panel.

At times you will need to use the terminal to fix conflicts, but this should only happen if you are collaborating with git.

Many people use third-party apps, but these have their own learning curve.

The basic process for using Git as a single user is this:

* save all of your open files
* check the status of your repository so you know what files have changed
* add them to the Git index of files
* commit them to create a save point for all of the files in your project
* push the commit to the remote repository (github)

Here are the commands from ```git_commands.R```

```{r eval = FALSE}
# basic everyday commands for all git users -------------------------------
 
gert::git_status()
gert::git_add("*")
gert::git_commit("a short comment describing the changes made")
blaseRtemplates::git_push_all()

```

That is it!

If you run into problems, the error messages are usually informative.  If you have trouble understanding an error or it isn't helpful, just google it.

One common use-case for git for the single user is to "un-delete" files.  You can do this as long as the files were previously committed.  To go back in time and recover the previously committed version, first make a commit to create a save point and then in the terminal run ```git log``` to review your commit history.  Add and commit any outstanding changes to your repository as outlined above.  Then run ```git checkout <commit id>``` to rewind the state of your repository to the desired commit.  Copy the file(s) you want to resurrect to a location external to the repository. Run ```git switch -``` to go back to your latest commit.  Make your changes and go from there.

Here is a way that is even easier and doesn't require using the terminal:  go to your github site for that repo.  Click on clock icon labled "commits" (which helpfully includes a "back in time" arrow).  Then select from the list of commits (versions).  Click browse files and find what you were looking for.  You can even click on the raw version to copy and paste directly back into R studio.

Here's a more complicated situation most of us can relate to:

* You submit your beautiful manuscript to Journal A
* Journal A bounces it right back to you
* You reformat all of the figures for your paper in R and submit it to Journal B
* Journal B is more interested, and they send it out for review and 6 weeks later reject it.
* You decide to submit to Journal C which has similar figure formatting as does Journal A.
* What do you do? 

```{r}
# run these commands to rewind to a prior "good" commit ----------------------

# make sure git status is "clean" (all changes committed) before rewinding
# gert::git_log() #find the id of the good commit
# blaseRtemplates::git_rewind_to(commit = "<good commit id>")

```

Hopefully you have a commit in your git log with the message "submitted to Journal A".  You select that id (the long character string) and provide it to the ```blaseRtemplates::git_rewind_to()``` function.  This function will generate a new commit that reverses all of your changes from present, back to the indicated commit.  Nothing is lost or overwritten in the process of going back in time.  The only thing that happens is that inverse changes are applied.  So if you need to go to Journal D (which is formatted like Journal B) you can always do that later.  

## Exercises

1.  Use the code blocks from this vignette to make a new R project on your system and for extra credit push it to github.  
    * First, establish the blaseRtemplates installation
    * Next create a new project
    * Connect your system to your github account using the git_commands.R script
    * Last, run through the initialization script and delete it

2.  Explore the scripts in the R directory. 

3.  Try running the code in dependencies.R to install some recommended packages.

4.  For more extra credit:  save your work, then add, commit, and push to your github site.
