---
title: "Week 2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{week_2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include = FALSE}
library(datascience.curriculum)
```

## Introduction

Today we discuss fundamentals of working with data in R.  

There is always more than one way to accomplish the task of reading, analyzing and plotting data.  My perspective is that this should be accomplished with computer code that is easy to read and error-resistant.  In my opinion, the "tidyverse" family of packages achieve this goal moreso than "base" R functions or other packages a large percentage of the time.  So I will mostly show you the tidyverse way.

## Data formatting

The best way to make the coding easy is to start by formatting your data in a proper way.

There is [theory](https://en.wikipedia.org/wiki/Database_normalization) on this if you are interested.  In short your data should:

* be in a flat text file format, such as .csv (comma separated values), .tsv (tab separated values) or .txt (plain text).  The file extensions don't matter for R; they only affect which program reads your file on mac or PC.  Unlike excel files, these files can be tracked by git so you know if any changes have been made accidentally.
* each row represents an observation
* each column represents a variable or attribute for that observation
* each column should be a single specific "type":  numeric, integer, factor and character are the most common types we use   
* each row has a unique "key".  The key can be a single column or a combination of columns as long as it distinguishes each row from every other row.

## Input and data types

Data should exist as a fixed, unchanging file on your hard drive.  You want to read it into R using ```readr::read_csv()```.  This is a tidyverse command that reads the file, automatically infers data types, and formats the data as a "tibble".  A tibble is related to the dataframe from base R but has superior characteristics for almost everything we do.  Think of it as a programmatic representation of your csv file within R.

```{r}
# attach the packages we will need
library(tidyverse)
# read in an example data file
# or substitute a file path to your own data
read_csv(file = "https://raw.githubusercontent.com/blaserlab/datascience.curriculum/main/inst/extdata/demo_iris_data.csv")
```

These data are from a collection of sample datasets provided with R.  As provided, it lacks a useful "key" column, so I added the observation column in the first position.  The data indicate measurements for various species of iris flower.

When read into R, the read_csv command infers the data types from the values provided and provides a message for you explaining what it has done.  In this case, they are all correct except for observation.  Why?  Because it has interpreted observation as "dbl" which is another way of saying "numeric" which in this case is not really correct.  Observation is just a serial number of the observations which in this case have an inherent order but not a quantitative value.  So we can tell R to interpret this as a factor which is the R data type for ordered data.


```{r}
read_csv(file = "https://raw.githubusercontent.com/blaserlab/datascience.curriculum/main/inst/extdata/demo_iris_data.csv", col_types = "fddddc")
```

The string "fddddc" tells R to interpret the columns as "factor", "double" or "character".

**You should always check that the data types are correct when you import your data in order to prevent errors.**

### More on data types

* character:  any string of valid alphanumeric characters.  Carries no order or quantitative value
* numeric/double:  any real number
* integer:  any integer, a subset of the real numbers
* factor:  factors are essentially characters bound to integer values. The character part is called a "level".  Factors can be a little tricky but are useful for ordering and reordering categorical data.

## Variable assignment

An important advantage of using a programming language to analyze data is the concept of abstraction, or assigning complicated values to simple variable names.

```{r}
# assign the data to the variable, demo_data
# readr::read_csv(file = system.file("extdata/demo_iris_data.csv", package = "datascience.curriculum"))
demo_data <- read_csv(file = "https://raw.githubusercontent.com/blaserlab/datascience.curriculum/main/inst/extdata/demo_iris_data.csv", col_types = "fddddc")
demo_data
```

The left arrow operator stores the data in the global environment under the name "demo_data". 

Any valid R object (data, functions etc) can be stored under a variable name.  By default they will be stored in the global environment.  If you save something else with the same name in the same environment it will overwrite the old value without checking, asking or notifying you.

## Data structures

Dataframes and tibbles are essentially lists of vectors with some rules attached. So what are lists and vectors?

A vector is the fundamental data structure in R.  It is a collection of data values, all of which must be the same data type.

You can make your own vector directly in R like so:

```{r}
# a numeric vector
c(1, 2, 3)

# a character vector
c("a", "b", "c")

# another character vector
c("1", "b" ,"charlie")

# here 1 gets coerced to a character because a vector must be all the same type
c(1, "b", "charlie")
```

In a dataframe or a tibble, each column is a vector. 

We can use some R operators to extract the columns as vectors.

```{r}
# extract a column by position
# the head command prints the first few values only
# omit head() if you want the whole thing
head(demo_data[[2]])

# extract a column by name
head(demo_data[["Species"]])

# another way to extract a column by name
head(demo_data$Species)

```

Usually you want to use the ```$``` operator to extract a column by name.  

A list is a more general form of data object.  It can hold any combination of R data types.  This is how you make a list:

```{r}
# make the list
demo_list <- list(1, "b", "charlie")

demo_list

# optionally name the elements
names(demo_list) <- c("a_number", "a_letter", "a_name")

# extract the elements by name
demo_list$a_number
demo_list$a_letter
demo_list$a_name

```

Lists are very flexible and useful objects for holding your data.

## Subsetting with base R functions

I think this syntax is suboptimal compared to Dplyr syntax, but for some advanced data objects you need to know it.  

For subsetting you use a single bracket rather than the double bracket for extracting data.  This returns an object of the same class, only smaller.  

```{r}
# subset using dataframe[row,colum] syntax
# subset a dataframe to get the second and third columns only
demo_data[,2:3]

# subset a dataframe to get the first and second rows only
demo_data[1:2,]

# subset to include columns by name 
demo_data[c("Species", "observation")]

# subset a list to return a smaller list
# return the first two elements
demo_list[1:2]

# subset a list by name
demo_list["a_name"]

```


## Data operations

In R the fundamental data structure is a vector.  A single value is a vector of length 1.  Many functions are optimized to work on the entire vector at once.

```{r warning=TRUE}
# get the mean of the sepal length
mean(demo_data$Sepal.Length)

# add two vectors
c(1, 2) + c(2, 3)

# for vectors of unequal length, the smaller vector is "recycled" for each element of the larger vector
1 + c(1, 2, 3)

# this gives a warning if the recycling doesn't work out evenly
c(1, 2) + c(1, 2, 3)

```

You can read more about arithmetic and logical operators [here](https://www.statmethods.net/management/operators.html).

Logical operators return a value of TRUE or FALSE.

```{r}
1 > 0

2 == 2

1 != 3

2 <= 1

```

## Dplyr 

Dplyr is a part of the tidyverse packages.  This is what you will use for data transformations prior to statistical testing or plotting.

Dplyr functions are pipe-friendly, meaning that they can be chained in a way that reduces redundant text in your code or saving intermediate values which can cause problems.  For this you use the pipe operator ```%>%``` or ```|>```.  The latter is a new version which I find preferable but requires R version 4.1 or later.  Either will work for all of the code we demonstrate here. 

Dplyr functions are also vectorized so they are computationally efficient and can be used on tables with hundreds of thousands of rows.

```{r}
# add a new column with mutate
# then group by a useful categorical variable
# then summarize the new value we calculated by mean according to group
demo_data |>
  mutate(sepal_l_w = Sepal.Length + Sepal.Width) |>
  group_by(Species) |>
  summarise(mean_sepal_l_w = mean(sepal_l_w))

```

You can use any built-in or custom function in mutate or summarise.  Functions for mutate must return a vector of the same length as the input vector; functions for summarise must return a vector of length 1. 

Sometimes you may find it useful to select or de-select columns from a tibble:

```{r}
# return a tibble without the observation column
demo_data |>
  select(-observation)
# return a tibble with only the Species and Sepal.Length columns
demo_data |>
  select(c(Species, Sepal.Length))
```

Use ```filter``` to select rows based on conditional test(s):

```{r}
# filter rows satisfying a logical test
demo_data |>
  filter(Species == "setosa")

demo_data |>
  filter(Species %in% c("setosa", "versicolor"))

```

In some cases it may be useful to have the data in "long form" as opposed to "wide form". Here you use ```pivot_longer``` and you are going to tell R which columns to convert to long form.  These should all be of the same type because they will end up in the same column.  The selected column names get repeated in a "name" column and the values end up in a "value" column.  The columns you don't select will be retained as additional attributes for the values. 

```{r}
# pivot from wide form to long form
long_data <- demo_data |>
  pivot_longer(cols = c(Sepal.Length,  Sepal.Width, Petal.Length, Petal.Width))
long_data

# pivot back to wide form
long_data |>
  pivot_wider(names_from = "name", values_from = "value")

```

Sometimes you have two tibbles with related information you want to join together.  This is a powerful tool for adding new attributes onto existing data.  

```{r}
# make two smaller tables
sepal_data <-
  demo_data |>
  group_by(Species) |>
  summarise(mean_sepal_l = mean(Sepal.Length), mean_sepal_w = mean(Sepal.Width))
sepal_data

petal_data <-
  demo_data |>
  group_by(Species) |>
  summarise(mean_petal_l = mean(Petal.Length), mean_petal_w = mean(Petal.Width))
petal_data

# now join them back together
left_join(sepal_data, petal_data)

```

Some rules for joining:

* you must have at least 1 column with common values to join by.  It is better if you have 1 and only 1.  The easiest way to do this is make sure that this column has the same name in both tables.
* you should avoid duplicated values in either of the tables you want to join.  Usually you want to keep the "main" data table the same and add on new data.  If the new data has duplicate entries it will be duplicated in the output which may cause problems. 
* missing data will be filled in with NA which is usually OK.  The order of the data in the two tables doesn't matter.  For ```left_join(x, y)``` the order of the data in the resulting tibble will be the same as for x with the exception of possible duplications.
* always check to be sure your join worked OK

For more information on these operations see this useful [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).

## Exercises

* try running the example code here in your own project
* take data from your own experiments, reformat by hand in a csv if necessary and read into R
* explore different data manipulations with base R and dplyr functions
* try comparing groups using simple statistical tests like ```t.test()``` and ```wilcox.test()```.  Hint:  you just need to feed numeric vectors with your data into these functions.
